Massively Parallel Pseudo-Spectral DNS and~LES for Particle-Laden Turbulent Flows under Two-Way Momentum Coupling

### 2022-12-04 _base-text/msc-thesis.tex
initial fixes after azaba review

In TWC approach, we assume that cumulative force per unit mass exerted by particles on~the~fluid, denoted by $\mathbf{f}^{(p)}$ in Equation \ref{eqn:n-s}, is non-zero.
Such contribution may be expressed as follows \parencite{Bosse2006,Monchaux2017,Rosa2022}:
\begin{equation}
\mathbf{f}^{(p)}(\mathbf{x}, t) = - \frac{M}{\rho_{p}} \sum_{k=1}^{N_{\text{part}}} m^{(k)}_{p} \left( \frac{ \mathbf{U}(\mathbf{Y}^{(k)}(t), t) - \mathbf{V}^{(k)}(t) }{\tau_{p}} \right) \delta \left( \mathbf{x} - \mathbf{V}^{(k)}(t) \right) ,
\label{eqn:twc-th}
\end{equation}
where most of the notation follows that from Equation \ref{eqn:owc}, except $m^{(k)}_{p}$ which denotes individual mass of $k$-th particle.
Moreover, $M$ is the weighting (super-particle) factor, discussed later in this section.
This continuous description involves summation of contributions from all particles (terms that, according to Newton's third law, are equal in magnitude and opposite in direction to terms in Equation \ref{eqn:owc}), and localised in given spatial point $\mathbf{x}$ with Dirac delta distribution.
In practice, we need a convenient way to calculate $\mathbf{f}^{(p)}$ for positions $\mathbf{x}$ at grid points without need to iterate over all particles.
Due to mismatch between actual particle positions and discrete locations of grid points we use mollified Dirac delta, so that strict particle locality is relaxed and their contributions may be projected onto adjacent grid nodes.
Thus,~Equation \ref{eqn:twc-th} becomes
\begin{equation}
\mathbf{f}^{(p)}(\mathbf{x}, t) = - \phi_{m}(\mathbf{x}, t) \left( \frac{ \mathbf{U}(\mathbf{x}, t) - \mathbf{V}_{E}(\mathbf{x}, t)}{\tau_{p}} \right) ,
\label{eqn:twc-px}
\end{equation}  
where $\phi_{m}$ is local mass loading. 
The term $\mathbf{V}_{E}(\mathbf{x}, t)$, is referred to as Eulerian particle velocity at location $\mathbf{x}$, and may be computed as follows:
\begin{equation}
\mathbf{V}_{E}(\mathbf{x}, t) = \sum_{k=1}^{N_{\text{part}}} \mathbf{V}^{(k)}(t) \, \sigma(|| \mathbf{x} - \mathbf{Y}^{(k)}(t) ||) ,
\label{eqn:twc-ve}
\end{equation}
where $\sigma$ is the interpolation kernel used to determine effect of given particle, based on its location, $\mathbf{Y}^{(k)}(t)$, onto grid node located at $\mathbf{x}$.
There are several choices of such kernels and their practical implementations \parencite{Garg2007}, but in this study the \emph{projection onto neighbouring nodes} (PNN) is used exclusively.
Instead of summing all particle forces around a node, particle force contribution is projected onto $8$ neighbouring nodes based on some weighing scheme.
Such scheme, in general, is a standard bi- or tri-linear function, that depends on separation distance between particle and grid node.
Here, in particular, we~use calculations based on cell volume partition (see Figure \ref{fig:pnn}) introduced in \textcite{Squires1990}.
This interpolation procedure is justified when influence of particles on the fluid is~indeed highly localised and limited to the closest neighbouring nodes.
In practice, this takes form of requirements that size (radius, $a$) of a particle must be smaller than both spatial grid spacing, $\Delta x$, and smallest resolved turbulent scales, as measured by Kolmogorov length scale, $\eta$.
Both conditions are satisfied for all simulations in this study (see Section \ref{ssc:ch2.flow.part}).

@@

\begin{figure}[h]
\centering
\includegraphics[width=8cm]{img/plots/2-1-2a-spec.pdf}
\caption{
The normalised energy spectra of background turbulent flows for both DNS and LES. Dashed line provides comparison with $k^{-5/3}$ curve of theoretical Kolmogorov spectrum (see Equation \ref{eqn:kolm}) for inertial subrange.
}
\label{fig:spec}
\end{figure}

@@

Results of LES show promising alignment with those from DNS.
It is important to recall that energy and dissipation spectra are directly available for smaller range of wavenumbers in LES, which coincides with part of DNS spectrum for, approximately $k \eta < 0.25$.
For non-settling particles, this is the part of energy and dissipation spectra in DNS that decrease with increasing mass loading, and it is also reflected in results for LES.
Similarly, when effects of gravity are included, sharp increase in both spectra throughout all modes (for $|\mathbf{k}| > 2$) is~witnessed, as expected.

The dual influence of TWC on energy spectrum for non-settling particles is known as ''pivoting'' \parencite{Squires1990,Bosse2006,Rosa2020}.
The kinetic energy of fluid is suppressed for intermediate wavenumbers due to incresed energy dissipation, while it is increased for high wavenumbers due to combined effects of larger viscous dissipation and transfer of momentum from the particles to the fluid.
The increase of energy for small turbulent scales cannot be adequately represented in LES, in part due to the fact that used subgrid-scale model is agnostic to any momentum transfer form particles. 
Similar remarks are valid for the settling particles, but here increase due to transfer of momentum from particles is larger and present throughout wide range of wavenumbers, and such transfer affects even intermediate turbulent scales that are directly resolved in LES.

@@

We can easily see that both statistics are increasing with $\Phi_m$ for settling particles, while in the absence of gravity the effect is opposite.
Also trends in both $u'$ and $\epsilon$ is somewhat expected, as these statistics are somewhat related to the total kinetic energy of turbulence ($u' \propto \sqrt{E}$).
In particular, for settling particles, the steep increase of both statistics with particle mass loading may be explained by larger fluid velocity gradients caused by the fast settling particles.
Also note, that due to different way of plotting results (relative to $\Phi_m$ and not number of particles), we may also observe that $u'$ is much more dependent on radii of individual droplets, while $\epsilon$ is much less sensitive to that, and seems to be mainly driven by the particle mass loading in the system.
This difference may be seen in results for both DNS and LES.

@@

When gravity is not considered, the results of LES significantly underestimate the RDF, including peak values around $St=1$ being reduced by approximately $60 \%$ \parencite{Fede2006,Yang2008,Marchioli2008}.
This difference may be attributed to filtering of~small-scale turbulence which has significant impact on concentration of particles of that size.
On the other hand, LES is confirmed \parencite[see]{Rosa2017} to be much more accurate in predicting the RDF of settling particles.
In general, it is argued that heavy settling particles due to increased velocity in the direction of gravity vector spend less time interacting with turbulent eddies and more uniformly sample the flow \parencite{Ireland2016}.
This manifests, in general, in reduced clustering due to turbulence (lower RDF) for settling particles, irrespective of method used.
This also explains why LES is much more accurate in~this~case, as filtered scales of turbulence have much more limited impact on particles, then in case of no gravity.

@@

In case of the relative radial velocity, methodology used in its computation is similar to~the~RDF, since comparable power-law relation also exists \parencite[Equation 13]{Rosa2013}.
For any two particles with separation vector $\mathbf{r}$ and relative velocity $\mathbf{w}$ we define their \emph{radial relative velocity} as:
\begin{equation}
w_r = \mathbf{w} \cdot \frac{\mathbf{r}}{|\mathbf{r}|} ,
\label{eqn:rrv}
\end{equation} 
i.e. dot product of relative velocity with direction vector of their separation.
Note, that $w_r$ is largest for two particles speeding towards each other.
Relative velocities are averaged within shells and the fitted into exponential function in similar fashion as with the RDF.
Then, they are averaged over time, to obtain $\langle|w_{r, 11}(r = R)|\rangle$, and finally normalised using the Kolmogorov velocity scale, $v_k$, to make resulting value dimensionless.

@@

The collision kernel may be arduously computed by tracking all collision events in the system and normalizing it by the number of all possible particle pairs.
Such value is referred to as the \emph{dynamic collision kernel} and denoted by $\Gamma^D_{11}(t)$, or just $\Gamma^D_{11}$ when averaged over~time.
Statistical uncertainty of such process is calculated as in \textcite[Equation 17]{Rosa2013}.
An~alternative formulation is the so-called \emph{kinematic collision kernel}, $\Gamma^K_{11}{t}$, based on \textcite{Sundaram1997}, where it is defined as a function of kinematic statistics computed in~previous section, and expressed by:
\begin{equation}
\Gamma^K_{11} = 2 \pi R^{2} \langle | w_{r, 11}(r = R) | \rangle g_{11}(r = R),
\label{eqn:gamma-k}
\end{equation}
where $r$ is the separation distance between particles and $R = 2a$.
Note that values or both kernels are expected to be in almost perfect agreement (up to statistical uncertainties), and presented simulation result confirm that expectation.

@@

The results are presented for both kinematic statistics (RDF, RRV) and collision kernels.
Note that values of both dynamic and kinematic kernels were confirmed to be in quantitative agreement within statistical uncertainty, hence only data for dynamic kernel is shown.
Three types of plots are used to showcase influence of two-way momentum coupling on these statistics.
Firstly, they are plotted for selected fixed values of $\Phi_m$ depending on particle radii (see Figures \ref{fig:twcrdf}, \ref{fig:twcrrv}, and \ref{fig:twcgamma}; top row).
These may be considered an extension of plots from Sections \ref{ssc:ch2.coll.rdfrrv} and \ref{ssc:ch2.coll.gamma}, since they also contain similar curves shown for OWC simulations, extended by additional ones that involve TWC with increasing mass loading.
Remaining plots show droplet statistics for two radii ($30$~$\upmu\text{m}$ and $40$~$\upmu\text{m}$), chosen to represent particles most affected by turbulence (i.e. with $St \sim 1$), depending on the amount of particles in the system.
Plots in the bottom row of Figures \ref{fig:twcrdf}, \ref{fig:twcrrv}, and \ref{fig:twcgamma} is limited to relatively small particle content (up to 20 million; equivalent to $\Phi_m \approx 0.15$ for $a = 40$~$\upmu\text{m}$).
The values obtained from OWC simulations are included for reference on the leftmost edge of the plot (i.e.~$0$~on~the~$X$-axis). 
Results from simulations with broader range of mass loadings are shown in Figures \ref{fig:twcrdf}, \ref{fig:twcrrv}, and \ref{fig:twcgamma}.
Simulations where super-particle parameterisation was used (i.e. $M>1$) are explicitly marked by dashed line.
For more information pertaining to simulations with super-particle parameterisation and what exact values of $M$ were used, see Appendix \ref{app:spp}.
Results without effects of gravity are shown on the left, and those for settling particles - on the right.

@@

The results of LES for non-settling particles are consistent with those of DNS in terms of~general trends and significant reduction of the RDF as observed in simulations with OWC.
As observed in \textcite{Rosa2020} the RDF is barely sensitive to TWC and larger mass loadings (even as large as $\Phi_m = 1$, shown here) for both very small ($20$~$\upmu\text{m}$) and very large particles ($60$~$\upmu\text{m}$).
It is so, because small inertia particles have little impact on turbulence modulation and their behaviour in dominated by the fluid flow, while heavy particles are not very sensitive to influence of turbulence in general, and turbulence modulated by other particles in particular.
The same conclusion is valid for results from LES.
One interesting difference, is the fact that results of LES are much less influenced by increasing mass loading.
For $\Phi_m \le 0.1$ mass loading seems to have very little effect on the RDF, and it can be seen for very large number of particles ($\Phi_m = 0.5, \, 1$), but still it remains larger that for DNS with corresponding mass loadings.
As shown before, see Section \ref{ssc:ch2.flow.twc}, large amount of heavy particles (when gravity is neglected) has tendency to weaken turbulence, which may result in reduced particle clustering.
In LES, smallest turbulent scales are parameterised by subgrid-scale model which is agnostic to any momentum transfer from particles to fluid, hence such dampening of smallest fluctuations cannot happen.
This may explain where, up to some point, LES is not sensitive to increasing mass loadings in TWC.
Still, large enough $\Phi_m$ are able to affect even larger turbulent scales that are directly resolved in LES, but this reduction is still not as dramatic as for DNS.

@@

Interestingly, as in the case of OWC (see Figure \ref{fig:owcrdfrrv}), the results obtained by LES are much more aligned with DNS for settling particles.
When gravity is not considered, the RDF remains close to constant and almost completely insensitive to the increase in mass loading.
We have to look at larger picture (Figure \ref{fig:twcrdfext}) to notice that results from LES become sensitive to $\Phi_m$ at higher loadings ($\sim 0.3$), where they start to gradually decrease.
This is consistent with previous observations and its explanation.
When settling particles are considered, the results of LES seem to follow those from DNS more closely.
In general, there is small overestimation on part of LES, but for the most part it remains within acceptable range.
These discrepancies seem to be more significant for lighter particles, as they are more susceptible to TWC-agnostic turbulence from subgrid-scale model, as well as less capable of~turbulence modulation due to filtering of small-scale turbulence.
For heavier particles ($40$~$\upmu\text{m}$ these differences are smaller in absolute value and show consistent overprediction.
Moreover, based on data from Figure \ref{fig:twcrdf} (top right), it is expected that for larger particles results for DNS and LES should be more and more aligned.

@@

When we analyse dependence of $\Gamma^D_{11}$ on the amount particles, again we can see that behaviour is different for non-settling and settling particles.
For the former, the collision kernel slightly decreases, while for the latter we can observe more definite increase in collision rate with increasing mass loading.
The collision kernel estimated by LES is in relatively good agreement with results obtained by DNS, and discrepancies seem to diminish with increasing mass loading of the dispersed phase.
What is interesting, even though kinematic statistics obtaied by DNS and LES were more aligned for particles of larger radius, $40$~$\upmu\text{m}$ in particular, for collision kernel we see the reverse.
Results for $a = 40$~$\upmu\text{m}$ are well aligned, in case of $a = 30$~$\upmu\text{m}$ we observe almost perfect agreement between two methods, both for small and large mass loadings.
This may be attributed to cancellations of certain effects specific to LES on kinematic statistics, that arise from filtering of small scales of turbulence, which cause underestimation of the RDF and overestimation of the RRV.

@@

These parameters, however, are only related to the fluid flow computations and do not impact the effort that is necessary to resolve particles and particle-fluid interactions in the system.
This is especially vital in simulations with two-way momentum coupling, where particle-to-fluid ratio is of great importance, thus necessitating simulation of large numbers of computational particles to be tracked.
This issue has been already hinted at previously, as this is main reason for introduction of the super-particle parametrisation to achieve systems with considerably larger mass loadings.
Furthermore, it must be observed that codes used for DNS and LES are almost identical, barring additional LES code for computation of viscosity contribution from subgrid-scale model (see Table \ref{tab:code-steps}).
This has some clear advantages, as it provides better method comparability and code maintainability.
On the other hand, the parallel implementation of DNS is focused on splitting computations that become much less burdensome in LES as computations pertaining to particles tend to quickly dominate computational costs, as will be shown in the following sections.

@@

The label ``Global'' refers to data exchange between multiple, and often all, processes and requires large degree of synchronisation between processes.
It is associated with MPI functions: \texttt{MPI\_BCAST}, \texttt{MPI\_REDUCE}, and \texttt{MPI\_ALLREDUCE}.
It is often involved in stages used in computation of statistics.
The postprocessing of turbulence statistics (Stage 7) requires global data of the entire fluid velocity field.
On the other hand, gathering of particle-pair statistics (Stage 5) is mostly local, requiring only information on neighbouring particles in halo regions.
The only global aspect of its interprocess communication comes into play at the end, when all data is transferred to the one process that is responsible for disk I/O~operations.
The~\texttt{MPI\_ALLREDUCE} calls in Stages 9 and 10 cover necessary summations over fluid energy spectrum in the Fourier space.
In the former case, it is necessary to calculate and~redistribute to all processes the spectral-eddy viscosity in LES.
For the last stage, such global synchronization is necessary due to application of large-scale forcing, which is implemented in a sequential manner.
Also, to a limited degree, the data reduction along single axis is necessary in the symmetrisation of fluid velocity after Fourier transform \parencite{Ayala2014}. 
Thus, we may observe that such synchronisation in LES-specific code should not negatively impact entire execution flow, as another global communication is imposed soon after for DNS as well.

@@

Figure \ref{fig:pfsowc} shows the dependence of execution time on particle radius for simulations under one-way momentum coupling, as well as without particles.
We observe that WCT per time step increases monotonically with particle radius, both in DNS and LES, when gravity is considered and without it.
Moreover, in case of simulations with settling particles, the~impact of larger droplet radii is much stronger.
In part, these results may be explained by the fact that size of the halo region used to compute two-point particle statistics becomes larger with particle radii.
This means that more data related to the state of neighbouring particles needs to be transferred to the host process.
This implementation detail may explain slight increase in computational cost, as in case of results without gravity, but hardly such increase as is observed when gravity is considered.
More importantly, the clear difference between performance of simulations with and without gravity, predominantly for high inertia droplets ($a > 40$~$\upmu\text{m}$), cannot be explained in simple terms, as enabling gravity in the solver code should have no direct bearing on its computational complexity
This effect shall be further examined in following sections.

@@

The above observation is thoroughly confirmed by similar measurements made for simulations under two-way momentum coupling with varying number of particles (considering gravity).
These results are shown in Figure \ref{fig:pfstwc}.
When the number of particles is increased, execution times per step quickly become prohibitively large.
That is the reason behind very limited range (up to 5 million) of particle counts in LES simulations, and any larger mass loadings had to be achieved via superparticle parametrisation.
This is in stark contrast with DNS, where simulations with 20 million computational particles could be easily performed.
For illustration, on the hardware used to run these simulations (\emph{Okeanos}), the total running time of LES with more than 5 million particles would easily exceed 2 days (i.e. maximal time of a single job), while DNS with 20 million particles (largest used) would be able to compute almost $2.5$ times more time steps in the same amount of wallclock time.
Longer simulation times are beneficial, as they result in lower statistical uncertainty of the modelled parameters.

>>>>

In TWC approach, we assume that cumulative force per unit mass exerted by particles on~the~fluid, denoted by $\mathbf{f}^{(p)}$ in Equation \ref{eqn:n-s}, is non-zero.
Such contribution may be expressed as follows \parencite{Bosse2006,Monchaux2017,Rosa2022}:
\begin{equation}
\mathbf{f}^{(p)}(\mathbf{x}, t) = - \frac{M}{\rho_{p}} \sum_{k=1}^{N_{\text{part}}} m^{(k)}_{p} \left( \frac{ \mathbf{U}(\mathbf{Y}^{(k)}(t), t) - \mathbf{V}^{(k)}(t) }{\tau_{p}} \right) \delta \left( \mathbf{x} - \mathbf{Y}^{(k)}(t) \right) ,
\label{eqn:twc-th}
\end{equation}
where most of the notation follows that from Equation \ref{eqn:owc}, except $m^{(k)}_{p}$ which denotes individual mass of $k$-th particle.
Moreover, $M$ is the weighting (super-particle) factor, discussed later in this section.
This continuous description involves summation of contributions from all particles (terms that, according to Newton's third law, are equal in magnitude and opposite in direction to terms in Equation \ref{eqn:owc}), and localised in given spatial point $\mathbf{x}$ with Dirac delta distribution.
In practice, we need a convenient way to calculate $\mathbf{f}^{(p)}$ for positions $\mathbf{x}$ at grid points without need to iterate over all particles.
Due to mismatch between actual particle positions and discrete locations of grid points we use mollified Dirac delta, so that strict particle locality is relaxed and their contributions may be projected onto adjacent grid nodes.
Thus,~Equation \ref{eqn:twc-th} becomes
\begin{equation}
\mathbf{f}^{(p)}(\mathbf{x}, t) = - \phi_{m}(\mathbf{x}, t) \left( \frac{ \mathbf{U}(\mathbf{x}, t) - \mathbf{V}_{E}(\mathbf{x}, t)}{\tau_{p}} \right) ,
\label{eqn:twc-px}
\end{equation}  
where $\phi_{m}$ is local mass loading. 
The term $\mathbf{V}_{E}(\mathbf{x}, t)$, is referred to as Eulerian particle velocity at location $\mathbf{x}$, and may be computed as follows:
\begin{equation}
\mathbf{V}_{E}(\mathbf{x}, t) = \sum_{k=1}^{N_{\text{part}}} \mathbf{V}^{(k)}(t) \, \sigma(|| \mathbf{x} - \mathbf{Y}^{(k)}(t) ||) ,
\label{eqn:twc-ve}
\end{equation}
where $\sigma$ is the interpolation kernel used to determine effect of given particle, based on its location, $\mathbf{Y}^{(k)}(t)$, onto grid node located at $\mathbf{x}$.
There are several choices of such kernels and their practical implementations \parencite{Garg2007}, but in this study the \emph{projection onto neighbouring nodes} (PNN) is used exclusively.
Instead of summing all particle forces around a node, particle force contribution is projected onto $8$ neighbouring nodes based on some weighing scheme.
Such scheme, in general, is a standard bi- or tri-linear function, that depends on separation distance between particle and grid node.
Here, in particular, we~use calculations based on cell volume partition (see Figure \ref{fig:pnn}) introduced in \textcite{Squires1990}.
This interpolation procedure is justified when influence of particles on the fluid is~indeed highly localised and limited to the closest neighbouring nodes.
In practice, this takes form of requirements that size (radius, $a$) of a particle must be smaller than both spatial grid spacing, $\Delta x$, and smallest resolved turbulent scales, as measured by Kolmogorov length scale, $\eta$.
Both conditions are satisfied for all simulations in this study (see Section \ref{ssc:ch2.flow.part}).

@@

\begin{figure}[h]
\centering
\includegraphics[width=8cm]{img/plots/2-1-2a-spec.pdf}
\caption{
The normalised energy spectra of background turbulent flows for both DNS and LES. Dashed line provides comparison with $k^{-5/3}$ curve of theoretical Kolmogorov spectrum (see Equation~\ref{eqn:kolm}) for inertial subrange.
}
\label{fig:spec}
\end{figure}

@@

Results of LES show promising alignment with those from DNS.
It is important to recall that energy and dissipation spectra are directly available for smaller range of wavenumbers in LES, which coincides with part of DNS spectrum for approximately $k \eta < 0.25$.
For non-settling particles, this is the part of energy and dissipation spectra in DNS that decrease with increasing mass loading, and it is also reflected in results for LES.
Similarly, when effects of gravity are included, sharp increase in both spectra throughout all modes (for $|\mathbf{k}| > 2$) is~witnessed, as expected.

The dual influence of TWC on energy spectrum for non-settling particles is known as ''pivoting'' \parencite{Squires1990,Bosse2006,Rosa2020}.
The kinetic energy of fluid is suppressed for intermediate wavenumbers due to increased energy dissipation, while it is increased for high wavenumbers due to combined effects of larger viscous dissipation and transfer of momentum from the particles to the fluid.
The increase of energy for small turbulent scales cannot be adequately represented in LES, in part due to the fact that used subgrid-scale model is agnostic to any momentum transfer form particles. 
Similar remarks are valid for the settling particles, but here increase due to transfer of momentum from particles is larger and present throughout wide range of wavenumbers, and such transfer affects even intermediate turbulent scales that are directly resolved in LES.

@@

We can easily see that both statistics are increasing with $\Phi_m$ for settling particles, while in the absence of gravity the effect is opposite.
Also trends in $u'$ and $\epsilon$ are somewhat expected, as both of these statistics are related to the total kinetic energy of turbulence ($u' \propto \sqrt{E}$).
In particular, for settling particles, the steep increase of both statistics with particle mass loading may be explained by larger fluid velocity gradients caused by the fast settling particles.
Also note, that due to different way of plotting results (relative to $\Phi_m$ and not number of particles), we may also observe that $u'$ is much more dependent on radii of individual droplets, while $\epsilon$ is much less sensitive to that, and seems to be mainly driven by the particle mass loading in the system.
This difference may be seen in results for both DNS and LES.

@@

When gravity is not considered, the results of LES significantly underestimate the RDF, including peak values around $St=1$ being reduced by approximately $60 \%$ \parencite{Fede2006,Yang2008,Marchioli2008}.
This difference may be attributed to filtering of~small-scale turbulence which has significant impact on concentration of particles of that size.
On the other hand, LES is confirmed \parencite[see][]{Rosa2017} to be much more accurate in predicting the RDF of settling particles.
In general, it is argued that heavy settling particles due to increased velocity in the direction of gravity vector spend less time interacting with turbulent eddies and more uniformly sample the flow \parencite{Ireland2016}.
This manifests, in general, in reduced clustering due to turbulence (lower RDF) for settling particles, irrespective of method used.
This also explains why LES is much more accurate in~this~case, as filtered scales of turbulence have much more limited impact on particles, then in case of no gravity.

@@

In case of the relative radial velocity, methodology used in its computation is similar to~the~RDF, since comparable power-law relation also exists \parencite[Equation 13]{Rosa2013}.
For any two particles with separation vector $\mathbf{r}$ and relative velocity $\mathbf{w}$ we define their \emph{radial relative velocity} as:
\begin{equation}
w_r = \mathbf{w} \cdot \frac{\mathbf{r}}{|\mathbf{r}|} ,
\label{eqn:rrv}
\end{equation} 
i.e. dot product of relative velocity with direction vector of their separation.
Note, that $w_r$ is largest for two particles speeding towards each other.
Relative velocities are averaged within shells and fitted into exponential function in a similar fashion as with the RDF.
Then, they are averaged over time, to obtain $\langle|w_{r, 11}(r = R)|\rangle$, and finally normalised using the Kolmogorov velocity scale, $v_k$, to make resulting value dimensionless.

@@

The collision kernel may be arduously computed by tracking all collision events in the system and normalising it by the number of all possible particle pairs.
Such value is referred to as the \emph{dynamic collision kernel} and denoted by $\Gamma^D_{11}(t)$, or just $\Gamma^D_{11}$ when averaged over~time.
Statistical uncertainty of such process is calculated as in \textcite[Equation 17]{Rosa2013}.
An~alternative formulation is the so-called \emph{kinematic collision kernel}, $\Gamma^K_{11}{t}$, based on \textcite{Sundaram1997}, where it is defined as a function of kinematic statistics computed in~previous section, and expressed by:
\begin{equation}
\Gamma^K_{11} = 2 \pi R^{2} \langle | w_{r, 11}(r = R) | \rangle g_{11}(r = R),
\label{eqn:gamma-k}
\end{equation}
where $r$ is the separation distance between particles and $R = 2a$.
Note that values or both kernels are expected to be in almost perfect agreement (up to statistical uncertainties), and presented simulation result confirm that expectation.

@@

The results are presented for both kinematic statistics (RDF, RRV) and collision kernels.
Note that values of both dynamic and kinematic kernels were confirmed to be in quantitative agreement within statistical uncertainty, hence only data for dynamic kernel is shown.
Three types of plots are used to showcase influence of two-way momentum coupling on these statistics.
Firstly, they are plotted for selected fixed values of $\Phi_m$ depending on particle radii (see Figures \ref{fig:twcrdf}, \ref{fig:twcrrv}, and \ref{fig:twcgamma}; top row).
These may be considered an extension of plots from Sections \ref{ssc:ch2.coll.rdfrrv} and \ref{ssc:ch2.coll.gamma}, since they also contain similar curves shown for OWC simulations, extended by additional ones that involve TWC with increasing mass loading.
Remaining plots show droplet statistics for two radii ($30$~$\upmu\text{m}$ and $40$~$\upmu\text{m}$), chosen to represent particles most affected by turbulence (i.e. with $St \sim 1$), depending on the amount of particles in the system.
Plots in the bottom row of Figures \ref{fig:twcrdf}, \ref{fig:twcrrv}, and \ref{fig:twcgamma} are limited to relatively small particle content (up to 20 million; equivalent to $\Phi_m \approx 0.15$ for $a = 40$~$\upmu\text{m}$).
The values obtained from OWC simulations are included for reference on the leftmost edge of the plot (i.e.~$0$~on~the~$X$-axis). 
Results from simulations with broader range of mass loadings are shown in Figures \ref{fig:twcrdf}, \ref{fig:twcrrv}, and \ref{fig:twcgamma}.
Simulations where super-particle parameterisation was used (i.e. $M>1$) are explicitly marked by dashed line.
For more information pertaining to simulations with super-particle parameterisation and what exact values of $M$ were used, see Appendix \ref{app:spp}.
Results without effects of gravity are shown on the left, and those for settling particles - on the right.

@@

The results of LES for non-settling particles are consistent with those of DNS in terms of~general trends and significant reduction of the RDF as observed in simulations with OWC.
As observed in \textcite{Rosa2020} the RDF is barely sensitive to TWC and larger mass loadings (even as large as $\Phi_m = 1$, shown here) for both very small ($20$~$\upmu\text{m}$) and very large particles ($60$~$\upmu\text{m}$).
It is so, because small inertia particles have little impact on turbulence modulation and their behaviour in dominated by the fluid flow, while heavy particles are not very sensitive to influence of turbulence in general, and turbulence modulated by other particles in particular.
The same conclusion is valid for results from LES.
One interesting difference is the fact that results of LES are much less influenced by increasing mass loading.
For $\Phi_m \le 0.1$ mass loading seems to have very little effect on the RDF, and it can be seen for very large number of particles ($\Phi_m = 0.5, \, 1$), but still it remains larger that for DNS with corresponding mass loadings.
As shown before, see Section \ref{ssc:ch2.flow.twc}, large amount of heavy particles (when gravity is neglected) has tendency to weaken turbulence, which may result in reduced particle clustering.
In LES, smallest turbulent scales are parameterised by subgrid-scale model which is agnostic to any momentum transfer from particles to fluid, hence such dampening of smallest fluctuations cannot happen.
This may explain where, up to some point, LES is not sensitive to increasing mass loadings in TWC.
Still, large enough $\Phi_m$ are able to affect even larger turbulent scales that are directly resolved in LES, but this reduction is still not as dramatic as for DNS.

@@

Interestingly, as in the case of OWC (see Figure \ref{fig:owcrdfrrv}), the results obtained by LES are much more aligned with DNS for settling particles.
When gravity is not considered, the RDF remains close to constant and almost completely insensitive to the increase in mass loading.
We have to look at larger picture (Figure \ref{fig:twcrdfext}) to notice that results from LES become sensitive to $\Phi_m$ at higher loadings ($\sim 0.3$), where they start to gradually decrease.
This is consistent with previous observations and its explanation.
When settling particles are considered, the results of LES seem to follow those from DNS more closely.
In general, there is small overestimation on part of LES, but for the most part it remains within acceptable range.
These discrepancies seem to be more significant for lighter particles, as they are more susceptible to TWC-agnostic turbulence from subgrid-scale model, as well as less capable of~turbulence modulation due to filtering of small-scale turbulence.
For heavier particles ($40$~$\upmu\text{m}$) these differences are smaller in absolute value and show consistent overprediction.
Moreover, based on data from Figure \ref{fig:twcrdf} (top right), it is expected that for larger particles results for DNS and LES should be more and more aligned.

@@

When we analyse dependence of $\Gamma^D_{11}$ on the amount of particles, again we can see that behaviour is different for non-settling and settling particles.
For the former, the collision kernel slightly decreases, while for the latter we can observe more definite increase in collision rate with increasing mass loading.
The collision kernel estimated by LES is in relatively good agreement with results obtained by DNS, and discrepancies seem to diminish with increasing mass loading of the dispersed phase.
What is interesting, even though kinematic statistics obtaied by DNS and LES were more aligned for particles of larger radius, $40$~$\upmu\text{m}$ in particular, for collision kernel we see the reverse.
Results for $a = 40$~$\upmu\text{m}$ are well aligned, in case of $a = 30$~$\upmu\text{m}$ we observe almost perfect agreement between two methods, both for small and large mass loadings.
This may be attributed to cancellations of certain effects specific to LES on kinematic statistics, that arise from filtering of small scales of turbulence, which cause underestimation of the RDF and overestimation of the RRV.

@@

These parameters, however, are only related to the fluid flow computations and do not impact the effort that is necessary to resolve particles and particle-fluid interactions in the system.
This is especially vital in simulations with two-way momentum coupling, where particle-to-fluid ratio is of great importance, thus necessitating simulation of large numbers of computational particles to be tracked.
This issue has been already hinted at previously, as this is main reason for introduction of the super-particle parameterisation to achieve systems with considerably larger mass loadings.
Furthermore, it must be observed that codes used for DNS and LES are almost identical, barring additional LES code for computation of viscosity contribution from subgrid-scale model (see Table \ref{tab:code-steps}).
This has some clear advantages, as it provides better method comparability and code maintainability.
On the other hand, the parallel implementation of DNS is focused on splitting computations that become much less burdensome in LES as computations pertaining to particles tend to quickly dominate computational costs, as will be shown in the following sections.

@@

The label ``Global'' refers to data exchange between multiple, and often all, processes and requires large degree of synchronisation between processes.
It is associated with MPI functions: \texttt{MPI\_BCAST}, \texttt{MPI\_REDUCE}, and \texttt{MPI\_ALLREDUCE}.
It is often involved in stages used in computation of statistics.
The postprocessing of turbulence statistics (Stage 7) requires global data of the entire fluid velocity field.
On the other hand, gathering of particle-pair statistics (Stage 5) is mostly local, requiring only information on neighbouring particles in halo regions.
The only global aspect of its interprocess communication comes into play at the end, when all data is transferred to the one process that is responsible for disk I/O~operations.
The~\texttt{MPI\_ALLREDUCE} calls in Stages 9 and 10 cover necessary summations over fluid energy spectrum in the Fourier space.
In the former case, it is necessary to calculate and~redistribute to all processes the spectral-eddy viscosity in LES.
For the last stage, such global synchronisation is necessary due to application of large-scale forcing, which is implemented in a sequential manner.
Also, to a limited degree, the data reduction along single axis is necessary in the symmetrisation of fluid velocity after Fourier transform \parencite{Ayala2014}. 
Thus, we may observe that such synchronisation in LES-specific code should not negatively impact entire execution flow, as another global communication is imposed soon after for DNS as well.

@@

Figure \ref{fig:pfsowc} shows the dependence of execution time on particle radius for simulations under one-way momentum coupling, as well as without particles.
We observe that WCT per time step increases monotonically with particle radius, both in DNS and LES, when gravity is considered and without it.
Moreover, in case of simulations with settling particles, the~impact of larger droplet radii is much stronger.
In part, these results may be explained by the fact that size of the halo region used to compute two-point particle statistics becomes larger with particle radii.
This means that more data related to the state of neighbouring particles needs to be transferred to the host process.
This implementation detail may explain slight increase in computational cost, as in case of results without gravity, but hardly such increase as is observed when gravity is considered.
More importantly, the clear difference between performance of simulations with and without gravity, predominantly for high inertia droplets ($a > 40$~$\upmu\text{m}$), cannot be explained in simple terms, as enabling gravity in the solver code should have no direct bearing on its computational complexity.
This effect shall be further examined in following sections.

@@

The above observation is thoroughly confirmed by similar measurements made for simulations under two-way momentum coupling with varying number of particles (considering gravity).
These results are shown in Figure \ref{fig:pfstwc}.
When the number of particles is increased, execution times per step quickly become prohibitively large.
That is the reason behind very limited range (up to 5 million) of particle counts in LES simulations, and any larger mass loadings had to be achieved via superparticle parameterisation.
This is in stark contrast with DNS, where simulations with 20 million computational particles could be easily performed.
For illustration, on the hardware used to run these simulations (\emph{Okeanos}), the total running time of LES with more than 5 million particles would easily exceed 2 days (i.e. maximal time of a single job), while DNS with 20 million particles (largest used) would be able to compute almost $2.5$ times more time steps in the same amount of wallclock time.
Longer simulation times are beneficial, as they result in lower statistical uncertainty of the modelled parameters.

This result, even though expected, is much more severe than anticipated, and -- given this code and parallelisation architecture -- makes LES completely untenable for TWC simulations with even moderate mass loading.
One simple remedy for this issue is the use of super-particle model, as in this study, but it combines two types of parameterisations, i.e.~superparticles and subgrid-scale model, whose interaction is not well known, thus rendering veracity of~such~results fragile at best.
In the long run, the best solution would be establishing the parallelisation strategy for LES that relies not only on decomposition of grid used to compute fluid flow (as~with~current implementation, tailored for DNS) but also for particles, which clearly are the bottleneck.
